### 2025-11-14 sections/2-methods.tex
yet another pass

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of $\rho$-NN search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(\textbf{X})}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:rnn-search-complexity}
\end{gather}

@@

While ordinarily we would compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{centered at the query} of radius $\rho_k$ to the mean cardinality, $\hat{|\mathcal{C}|}$, of clusters in $\mathcal{Q}$ at a radius equal to the mean of their radii, $\hat{r}$.
Assuming that $q$ is from the same distribution as $\textbf{X}$, by Equation~\ref{eq:methods:lfd-original}, we have $\hat{d} = \frac{\log{}\frac{\hat{|\mathcal{C}|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$, and rearranging, $\frac{\hat{r}}{\rho_k} = ( \frac{\hat{|\mathcal{C}|}}{k} )^{\hat{d}^{-1}}$.
This simplifies the \textsc{LeafSearch} component to $k ( 1 + 2 \cdot ( \frac{\hat{|\mathcal{C}|}}{k} ) ^ {\hat{d}^{-1}} )^{\hat{d}}$.

@@

We begin by letting $\mathcal{Q}$ be a list of 3-tuples $(\mathcal{C}, \delta^{+}_{\mathcal{C}}, m)$, where $m$ is the multiplicity of $\mathcal{C}$ in $\mathcal{Q}$.
$\mathcal{Q}$ initially contains only $(\mathcal{R}, \delta^{+}_{\mathcal{R}}, |\mathcal{R}| - 1)$ while $\mathcal{H}$ initially contains only the center of $\mathcal{R}$.
We then use QuickSelect, modified to account for multiplicities and to reorder $\mathcal{Q}$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $\mathcal{Q}$ such that $\left| B(q, \tau) \right| + |\mathcal{H}| \geq k$.

>>>>

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of $\rho$-NN search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(\textbf{X})}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
where $\mathcal{N}_{\hat{r}}(\textbf{X})$ is the minimum number of clusters of radius $\hat{r}$ needed to cover $\textbf{X}$, and $\rho$ is the search radius.

@@

While ordinarily we would compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball of radius $\rho_k$ \textit{centered at the query} to the mean cardinality, $\hat{|\mathcal{C}|}$, of clusters in $\mathcal{Q}$ at a radius equal to the mean of their radii, $\hat{r}$.
Assuming that $q$ is from the same distribution as $\textbf{X}$, by Equation~\ref{eq:methods:lfd-original}, we have $\hat{d} = \frac{\log{}\frac{\hat{|\mathcal{C}|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$, and rearranging, $\frac{\hat{r}}{\rho_k} = ( \frac{\hat{|\mathcal{C}|}}{k} )^{\hat{d}^{-1}}$.
This simplifies the \textsc{LeafSearch} component to $k ( 1 + 2 \cdot ( \frac{\hat{|\mathcal{C}|}}{k} ) ^ {\hat{d}^{-1}} )^{\hat{d}}$.

@@

We begin by letting $\mathcal{Q}$ be a list of 3-tuples $(\mathcal{C}, \delta^{+}_{\mathcal{C}}, m)$, where $m$ is the multiplicity of $\mathcal{C}$ in $\mathcal{Q}$.
Initially, $\mathcal{Q}$ contains only $(\mathcal{R}, \delta^{+}_{\mathcal{R}}, |\mathcal{R}| - 1)$ and $\mathcal{H}$ contains only the center of $\mathcal{R}$.
We then use QuickSelect, modified to account for multiplicities and to reorder $\mathcal{Q}$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $\mathcal{Q}$ such that $\left| B(q, \tau) \right| + |\mathcal{H}| \geq k$.

### 2025-11-14 sections/3-datasets.tex
yet another pass

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axes. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD, indicating a manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process. It seems to have little manifold structure, if any, instead exhibiting behavior akin to the Random dataset across three separate radial scales, indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

>>>>

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axes. The x-axis is tree depth; the y-axis is LFD of clusters at that depth, computed using Equation~\ref{eq:methods:lfd-original} with $r_1 = r$ and $r_2 = \frac{r}{2}$ where $r$ is the radius of the cluster. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD, indicating a manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process. It seems to have little manifold structure, if any, instead exhibiting behavior akin to the Random dataset across three separate radial scales, as indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

### 2025-11-14 sections/4-results.tex
yet another pass

\subsection{Scaling Behavior and Recall}
\label{sec:results:scaling-behavior-and-recall}

@@

On Silva (Fig.~\ref{fig:results:silva-scaling}), all CAKES algorithms initially exhibit throughput that decreases approximately linearly with cardinality, but then levels off at higher cardinalities.
DFS is consistently the fastest algorithm on this dataset.

@@

For each of the vector datasets, Fig.~\ref{fig:results:indexing} reports the time taken for each algorithm to build its index assuming that the hyper-parameters are set to maximize recall.

@@

  \vskip -0.1in
  \subfloat[Sift for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/sift-indexing.png}
    \label{fig:results:sift-indexing}
  }
  \subfloat[Random dataset for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/random-indexing.png}
    \label{fig:results:random-indexing}
  }
  \caption{Indexing and tuning time.
  We report the time taken by various algorithms to build and tune its index on various datasets.
  For CAKES, this includes the time required to build the tree and find the fastest internal algorithm.
  All axes are on logarithmic scales.
  For all datasets, we see that the indexing time for CAKES increases roughly linearly with the cardinality of the dataset.
  HNSW and ANNOY are the slowest at indexing.
  FAISS-Flat does not actually build an index;
  it provides a na\"{i}ve linear search in the FAISS family of algorithms.
  It is included only to illustrate the base cost of simply having a dataset.}
  \label{fig:results:indexing}
\end{figure}

>>>>

@@

On Silva (Fig.~\ref{fig:results:silva-scaling}), the throughput of all CAKES algorithms initially decreases with cardinality, but then levels off at higher cardinalities.
DFS is consistently the fastest algorithm on this dataset.

@@

For each of the vector datasets, Fig.~\ref{fig:results:indexing} reports the time taken for each algorithm to build its index.

@@

  \vskip -0.1in
  \subfloat[Sift for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/sift-indexing.png}
    \label{fig:results:sift-indexing}
  }
  \subfloat[Random dataset for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/random-indexing.png}
    \label{fig:results:random-indexing}
  }
  \caption{Indexing and tuning time.
  We report the time taken by each algorithm to build and tune its index on various datasets.
  For CAKES, this includes the time required to build the tree and find the fastest internal algorithm.
  For HNSW, ANNOY, and FAISS-IVF, this excludes any hyper-parameter tuning time, including only the time taken to build an index assuming that the recall-maximizing hyper-parameters are already known.
  All axes are on logarithmic scales.
  For all datasets, we see that the indexing time for CAKES increases roughly linearly with the cardinality of the dataset.
  HNSW and ANNOY are the slowest at indexing.
  FAISS-Flat does not actually build an index;
  it provides a na\"{i}ve linear search in the FAISS family of algorithms.
  It is included only to illustrate the base cost of simply having a dataset.}
  \label{fig:results:indexing}
\end{figure}

### 2025-11-14 sections/5-discussion.tex
brought back Partition
revived availability line
yet another pass

This work could be extended with additional domain-specific distance functions, e.g.\ Wasserstein~\cite{vallender1974calculation} for probability distributions and Tanimoto~\cite{bajusz2015tanimoto} for molecular structures, requiring only a Rust implementation of the distance function itself to be compatible with CAKES.
Examining the geometric and topological properties of datasets that favor one CAKES algorithm over another would also be valuable.
% An examination of the geometric and topological properties of datasets that favor one CAKES algorithm over another would also be valuable.

CAKES is implemented in Rust and available under the MIT license at https://github.com/URI-ABD/clam.

>>>>

This work could be extended with additional domain-specific distance functions, e.g.\ Wasserstein~\cite{vallender1974calculation} for probability distributions and Tanimoto~\cite{bajusz2015tanimoto} for molecular structures, requiring only a Rust implementation of the distance function itself to be compatible with CAKES.
Examining the geometric and topological properties of datasets that favor one CAKES algorithm over another would also be valuable.
Development of approximate variants of CAKES algorithms would allow users to trade off recall for higher throughput.
Finally, the cluster tree could be extended to support dynamic insertion and deletion of points, enabling CAKES to be used in streaming applications.

We provide a Rust implementation of CAKES under the MIT license at \url{https://github.com/URI-ABD/clam}.
